Compositional Quantitative Resource Analysis of C Programs

### 2014-06-25 paper/main.tex
different changes

  This paper describes a novel quantitative resource analysis
  framework that tackles these challenges for C programs.  The
  analysis framework consists of three parts.  First, the runtime
  resource cost of programs is formalized with a parametric cost
  semantics for CompCert Clight.  Second, a quantitative Hoare logic
  that can be used for interactive bound development is implement and
  proved sound with respect to the cost semantics in the Coq Proof
  Assistant.  Third, an automatic amortized resource analysis for C
  programs that computes derivations in the quantitative Hoare logic
  is developed and implemented.  Both, the quantitative logic and the
  automatic amortized analysis are naturally compositional and can be
  combined to semi-automatically derive global resource bounds.

@@

Static analysis of quantitative properties of imperative programs is
an active research area and recent years have seen many innovations.
Notable tools that have been developed include
SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.
%
While the these tools can derive impressive results for realistic
software, there still exist shortcomings that hamper the application
of the existing techniques in practice.
\begin{itemize}
\item Analysis tools are black boxes that either deliver a result or fail
  without enabling \emph{user interaction} to manually or
  semi-automatically derive bounds for challenging parts of the
  program.
\item The computed bounds are often \emph{non-compositional} local
  bounds (for a single (nested) loop) that are difficult to combine to
  global whole program bounds.
\item Existing techniques often rely on complex external tools such as
  abstract interpretation based invariant
  generation~\cite{GulwaniMC09} or translation of the program into a
  term-rewriting system~\cite{BrockschmidtEFFG14, SinnZV14} without
  providing \emph{verifiable certificates} for the correctness of the
  derived bound.
\end{itemize}
%
While there has been much progress in static quantitative analysis,
Knuth correctly points out in a recent interview~\cite{KnuthInter}
that the state-of-the-art in formal quantitative methods still falls
short in comparison with semantic techniques.
%
\begin{quote}
  % Consider, for example, a computer program that sorts a list of
  % numbers into order. 
  Thanks to the work of Floyd, Hoare, and others,
  we have formal definitions of semantics, and tools by which we can
  verify that sorting is indeed always achieved. My job is to go
  beyond correctness, to an analysis of such things as the program's
  running time $[\cdots]$. I'm 100\% sure that my recurrence correctly
  describes the program's performance, and all of my colleagues agree
  with me that the recurrence is "obviously" valid. Yet I have no
  formal tools by which I can prove that my recurrence is right. I
  don't really understand my reasoning processes at all!
  \vspace{-3ex}
  \begin{flushright}
    -- Donald E. Knuth, 2014
  \end{flushright}
\end{quote}
%
%
In this paper, we present a novel resource analysis framework for C programs
that is based on a solid semantic foundation. It works automatically,
efficiently and precisely for many examples, can be used interactively
for manual bound derivation if the automation fails, and produces
verifiable certificates for the bounds.  Our contributions are as
follows.
%
\begin{enumerate}
\item We define a operational cost semantics for C programs that
  accounts a constant cost for each statement.  The cost is parametric
  in a cost metric and the semantics models cost of terminating and
  non-terminating executions.  A specific \emph{tick} function can be
  inserted in the source code to further customize the cost model.

@@

It is often necessary, when specifying a program, to relate certain
invariants in the pre- and postcondition.  A standard solution of
Hoare logics to this problem is \emph{logical variables} (reference XXX).  These
additional variables also mentioned in the literature as auxiliary
state are constant across the derivation.  For example, if we note
$Z$ such a logical variable, we can specify the function
$\code{double}()$ which doubles the global variable $x$ as
$$
  \htriple{z = Z}{\code{double}()}{x = 2 \cdot Z}.
$$

When formalizing Hoare logics in a proof assistant one can either
fully specify the syntax and semantics of assertions and hence get a
deep embedding, or use the assertions of the host theory to get a
shallow embedding.  For simplicity, we used the latter approach in our
development.  This choice makes it possible to have logical variables
almost for free: we can simply use the variable and binding mechanisms
of Coq, our host theory.  When an logical variable is needed we
introduce it using a universal quantifier of Coq before deriving triples
in the logic, so the Coq theorem for the above example would look like
%
\begin{lstlisting}
Theorem dbl_triple: forall Z,
  htriple ($\lambda\sigma$.$\sigma$(x) = Z) ($\code{double}()$) ($\lambda\sigma$.$\sigma$(x) = 2*Z).
\end{lstlisting}

However, this trick alone is often not sufficient when working with a
recursive function $f$.  In that case we apply the {\sc L:Extend} rule
of the logic to assume the specification of $f$ then proceed to prove
the function body with this induction hypothesis.  But in this process
it can be the case that the induction hypothesis needs to be used with
a different value of a logical variable.  To cope with this problem,
assumptions in the function context $\Delta$ are universally quantified
over logical variables and the {\sc L:Extend} rule once again leverages
the host proof assistant to require that the triple on $f$'s body is
proved for every possible logical variable $y$.

@@

\section{Polynomial Potential}

\paragraph{Index Sets}

Let $V$ be a set of variables.  An \emph{index} $I \in \ind(V)$ is a
family that maps two-element sets of variables to natural numbers,
that is,
$$
I = (i_{\{x,y\}})_{\{x,y\} \subseteq V} \; .
$$
%
We identify a family $I$ with the set $\{ (\{x,y\},i_{\{x,y\}})
\mid \{x,y\} \subseteq V\}$.

Let $\ind(V)$ denote the set of all such indices.  We write $\ind$
instead of $\ind(V)$ if the set of variables $V$ is fixed or obvious
from the context.
%
We assume that allways $0 \in V$ and sometimes write $i_x$ instead of $i_{\{x,0\}}$.

The \emph{degree} $\deg(I)$ of an index $I = (i_{\{x,y\}})_{\{x,y\}
  \subseteq V}$ is defined as
$$
\deg(I) = \sum_{\{x,y\} \in V} i_{\{x,y\}} \;.
$$
We define $\ind_k(V) = \{ I \mid I \in \ind(V) \text{ and } \deg(i) \leq k
\}$ to be the set of indices of degree at most $k$.

\paragraph{Resource Polynomials}

Let $V$ be a set of variables.  An index $I \in \ind(V)$ denotes a
\emph{base polynomial} $P_I : \states \to \N$ for $V$ that maps a
program state $H$ to product of binomial coefficients (a natural
number).  We define
$$
P_I(H) = \prod_{{\{x,y\}} \subseteq V} \binom {|H(x){-}H(y)|} {i_{\{x,y\}}} \; .
$$
%
A \emph{resource polynomial} $R$ for the variable set $V$ is a
non-negative linear combination of the base polynomials for $V$.

\paragraph{Potential Annotations}

A \emph{potential annotation} for the variable set $V$ is a family
$$Q = (q_I)_{I \in \ind(V)}$$ 
of non-negative rational numbers.  Such an annotation denotes the
resource polynomial $R_Q$ that is defined by
$$
R_Q(H) = \sum_{I \in \ind(V)} q_I \cdot P_I(H) \; .
$$
%
We say that the potential annotation $Q$ is of degree $k$ if $q_I = 0$
for $I \in \ind(V)$ with $\deg(I) > k$.

\paragraph{Additive Shifts}

Let $Q$ be a potential annotation for a variable set $V$ and let
$\{x,y\} \subseteq V$ be a two-element variable set.  The
\emph{additive shift} with respect to $\{x,y\}$ is a potential
annotation $\shift_{\{x,y\}}(Q) = (q'_I)_{I \in \ind(V)} $ for $V$
that is defined through
$$
q'_I = q_I + q_{I^{\{x,y\}{+}1}} \; .
$$
For an index $I = (i_{\{x,y\}})_{\{x,y\} \subseteq V}$ we use the
notation $I^{\{x,y\}{+}k}$ to denote the index
$(i'_{\{x,y\}})_{\{x,y\} \subseteq V}$ such that
$$
i'_{\{t,u\}} = \left\{
  \begin{array}{ll}
    i_{\{t,u\}} + k  & \text{if } \{t,u\} = \{x,y\} \\
    i_{\{t,u\}} & \text{otherwise}
  \end{array}
\right.
\;.
$$
%
The additive shift for natural numbers reflects the identity 
\begin{equation}
\label{eq:shift}
\sum_{0 {\leq} i \leq {k}} q_i \binom{n+1}{i} = \sum_{0 {\leq} i \leq {k}} (q_i{+}q_{i+1}) \binom{n}{i}
\end{equation}
where $q_{k+1} = 0$.  It is used in the effect system if the
difference $n+1$ between two variables $x,y$ decreases by one.

\begin{lemma} Let $V$ be a set of variables with $x,y \in V$ and let
  $H$ be a program state. Let $|H'(t) {-} H'(u)| = |H(t) {-} H(u)|$
  for $\{t,u\} \neq \{x,y\}$ and let $|H'(x) {-} H'(y)| = |H(x) {-}
  H(y)| - 1$.
  %
  If $Q' = \shift_{\{x,y\}}(Q)$ then $R_Q(H) = R_{Q'}(H')$.
\end{lemma}

We now study the effect of multiple simultaneous shifts.  Let $Q$ be a
resource annotation for a variable set $V$ and let $U_1,\ldots,U_n
\subseteq V$ with $|U_i| = 2$ for all $i$ and $U_i \neq U_j$ for $i
\neq j$ be pairwise distinct two-element variable sets.  The
simulations additive shift $\shift_{U_1,\ldots,U_n}(Q)$ of $Q$ with
respect to $U_1,\ldots,U_n$ is defined by
$$
\shift_{U_1,\ldots,U_n}(Q) = \shift_{U_1}( \cdots \shift_{U_n}(Q) \cdots ) \; .
$$
%
\begin{proposition}
  Let $V$ be a set of variables and let $U_1,\ldots,U_n$ be pairwise
  distinct two-element variable sets.  Let $|H'(x) {-} H'(y)| = |H(x)
  {-} H(y)|$ for $\{x,y\} \not\in \{U_1,\ldots,U_n\}$ and let $|H'(x)
  {-} H'(y)| = |H(x) {-} H(y)| - 1$ for $\{x,y\} \in
  \{U_1,\ldots,U_n\}$.
  %
  If $Q' = \shift_{U_1,\ldots,U_n}(Q)$ then $R_Q(H) = R_{Q'}(H')$.
\end{proposition}
%
As shown by the following lemma, the order in which the shifts for the
individual $U_i$ are applied is insignificant.
%
\begin{lemma}
  Let $\sigma : \{1,\ldots,n\} \to \{1,\ldots,n\}$ be a
  permutation. Then $\shift_{U_1,\ldots,U_n}(Q) =
  \shift_{U_{\sigma(1)},\ldots,U_{\sigma(n)}}(Q)$.
\end{lemma}
%
For reasons of efficiency in the constraint generation, we give a more
direct formula for the simultaneous shift.  Let $I \in \ind(V)$ and
let $U_1,\ldots,U_n$ be pairwise distinct two-element variable sets.
We define the index $I^{U_1,\ldots,U_n + k}$ as the family $(i'_{\{x,y\}})_{\{x,y\} \subseteq V}$ such that
$$
i'_{\{t,u\}} = \left\{
  \begin{array}{ll}
    i_{\{t,u\}} + k  & \text{if } \{t,u\} \in \{U_1,\ldots,U_n\} \\
    i_{\{t,u\}} & \text{otherwise}
  \end{array}
\right.  \;.
$$

%
\begin{lemma}
  Let $V$ be a variable set and let $U_1,\ldots,U_n$ be pairwise
  distinct two-element variable sets.
  %
  Let $Q = (q_I)_{I \in \ind(V)}$ be a resource annotation for
  $V$ and let $ Q' = (q'_I)_{I \in \ind(V)}$ where
  $$
  q'_I = \sum_{\{j_1,\ldots,j_m\} \subseteq \{1,\ldots,n\} } q_{I^{U_{j_1},\ldots,U_{j_m}+1}} \; .
  $$
  Then $Q' = \shift_{U_1,\ldots,U_n}(Q)$.
\end{lemma}

@@

\section{NOTES}

\subsection{Things to say in the intro}

\paragraph{Quote of Knuth}

\begin{quotation}
  There's even a fundamental gap in the foundations of my main
  mathematical specialty, the analysis of algorithms. Consider, for
  example, a computer program that sorts a list of numbers into
  order. Thanks to the work of Floyd, Hoare, and others, we have
  formal definitions of semantics, and tools by which we can verify
  that sorting is indeed always achieved. My job is to go beyond
  correctness, to an analysis of such things as the program's running
  time: I write down a recurrence, say, which is supposed to represent
  the average number of comparisons made by that program on random
  input data. I'm 100\% sure that my recurrence correctly describes
  the program's performance, and all of my colleagues agree with me
  that the recurrence is "obviously" valid. Yet I have no formal tools
  by which I can prove that my recurrence is right. I don't really
  understand my reasoning processes at all! My student Lyle Ramshaw
  began to create suitable foundations in his thesis (1979), but the
  problem seems inherently difficult. Nevertheless, I don't lose any
  sleep over this situation.
\end{quotation}
http://www.informit.com/articles/article.aspx?p=2213858

\paragraph{to do}

\begin{itemize}
\item describe schedulability analysis for two schedulers
\item include program logic from the stack paper for arbitrary resources
\item what operational semantics to use? 
\item verify automatic tool with program logic
\item derive some crazy bounds (polynomial? logarithmic?) with the logic
\item logic can also be used by automation: put linear constraints in the precondition
\end{itemize}

\paragraph{Advanteges over previous work}

\begin{itemize}
\item integrates well with hand-derived bounds in a program logic
\item have formal soundness proof with respect to cost semantics
\item can attach certificate to the code (proof-carrying code)
\item can deal with resources that \emph{become available} (e.g, freeing memory) in an execution
\item derive global bounds for the program as a function of the inputs (as opposed to local loop bounds)
\item can handle advance control flow such as break, return, mutual recursive functions
\item do not rely on abstract-interpretation based invariant generation
\item show that constant factors are very precise with respect to the cost semantics
\item demonstrate that the analysis is applicable in the formal verification of system code
\end{itemize}

>>>>

  This paper describes a novel quantitative resource analysis
  framework that tackles these challenges for C programs.  The
  analysis framework consists of three parts.  First, the runtime
  resource cost of programs is formalized with a parametric cost
  semantics for CompCert Clight.  Second, a quantitative Hoare logic
  that can be used for interactive bound development is implement and
  proved sound with respect to the cost semantics in the Coq Proof
  Assistant.  Third, an automatic amortized resource analysis for C
  programs that computes derivations in the quantitative Hoare logic
  is developed and implemented.  It is the first automatic amortized
  analysis that can derive bounds that depend on negative integers and
  differences between numbers.  Both, the quantitative logic and the
  automatic amortized analysis are naturally compositional and can be
  combined to semi-automatically derive global resource bounds.

@@

Static analysis of quantitative properties of imperative programs is
an active research area and recent years have seen many innovations.
Notable tools that have been developed include
SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.
%
While the these tools can derive impressive results for realistic
software, there still exist shortcomings that hamper the application
of the existing techniques in practice.
\begin{itemize}
\item Analysis tools are black boxes that either deliver a result or fail
  without enabling \emph{user interaction} to manually or
  semi-automatically derive bounds for challenging parts of the
  program.
\item The computed bounds are often \emph{non-compositional} local
  bounds (for a single (nested) loop) that are difficult to combine to
  global whole program bounds.
\item Existing techniques often rely on complex external tools such as
  abstract interpretation based invariant
  generation~\cite{GulwaniMC09} or translation of the program into a
  term-rewriting system~\cite{BrockschmidtEFFG14, SinnZV14} without
  providing \emph{verifiable certificates} for the correctness of the
  derived bound.
\end{itemize}
%
While there has been much progress in static quantitative analysis,
Knuth correctly points out in a recent interview~\cite{KnuthInter}
that the state-of-the-art in formal quantitative methods still falls
short in comparison with semantic techniques.
%
\begin{quote}
  % Consider, for example, a computer program that sorts a list of
  % numbers into order. 
  $[\cdots]$ Thanks to the work of Floyd, Hoare, and others,
  we have formal definitions of semantics, and tools by which we can
  verify that sorting is indeed always achieved. My job is to go
  beyond correctness, to an analysis of such things as the program's
  running time $[\cdots]$. I'm 100\% sure that my recurrence correctly
  describes the program's performance, and all of my colleagues agree
  with me that the recurrence is "obviously" valid. Yet I have no
  formal tools by which I can prove that my recurrence is right. I
  don't really understand my reasoning processes at all!
  \vspace{-3ex}
  \begin{flushright}
    -- Donald E. Knuth, 2014
  \end{flushright}
\end{quote}
%
The goal of this work is to develop a resource analysis framework for
C programs that is based on a solid semantic foundation.  The choice
of C is primarily motivated by our ongoing work on the formally
verified hypervisor kernel CertiKOS~\cite{GuVFSC11}.  CertiKOS is
mainly developed in C and is supposed to provide verified guaranties
on timing and memory usage.  Moreover, C is also a natural choice
because it is still the most widely used language for system
development; in particular in safety-critical embedded and real-time
systems where resource bound analyses are often required by regulatory
authorities~\cite{May2013}.

In a recent work, we formally verified a bound on the stack usage of
CertiKOS

In this article, we present the first resource analysis framework for
C that makes it possible to combine automatically and interactively
derived resource bounds.  It works automatically, efficiently and
precisely for many examples, can be used interactively for manual
bound derivation if the automation fails, and produces verifiable
certificates for the bounds.  Our contributions are as follows.
%
\begin{enumerate}
\item We define a operational cost semantics for C programs that
  accounts a constant cost for each statement.  The cost is parametric
  in a cost metric and the semantics models cost of terminating and
  non-terminating executions.  A specific \emph{tick} function can be
  inserted in the source code to further customize the cost model.

@@

It is often necessary to relate certain invariants in the pre- and
postcondition when specifying a program.  A standard solution of Hoare
logic is to use \emph{logical variables}~\cite{Kleymann99}.  These
additional variables (also called auxiliary state in the literature)
are constant across the derivation.  For example, if $Z$ is such
a logical variable, we can specify the function $\code{double}()$
which doubles the global variable $x$ as
$$
  \htriple{z = Z}{\code{double}()}{x = 2 \cdot Z}.
$$
%
When formalizing Hoare logics in a proof assistant one can either
fully specify the syntax and semantics of assertions and hence get a
deep embedding, or use the assertions of the host theory to get a
shallow embedding.  For simplicity, we used the latter approach in our
development.  This choice makes it possible to have logical variables
almost for free: we can simply use the variable and binding mechanisms
of Coq, our host theory.  When an logical variable is needed we
introduce it using a universal quantifier of Coq before deriving
triples in the logic, so the Coq theorem for the above example would
look as follows.
%
\begin{lstlisting}
Theorem dbl_triple: forall Z,
  htriple ($\lambda\sigma$.$\sigma$(x) = Z) ($\code{double}()$) ($\lambda\sigma$.$\sigma$(x) = 2*Z).
\end{lstlisting}
%
However, this trick alone is often not sufficient when working with a
recursive function $f$.  In that case we apply the {\sc L:Extend} rule
of the logic.  First we add a specification $\forall z\,\vec
v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)$ of $f$ to the function context
$\Delta$.  Then we proceed to prove the function $S_f$ body with this
induction hypothesis.  In this process it can be the case that the
induction hypothesis needs to be used with a different value of a
logical variable.  To cope with this problem, assumptions in the
function context $\Delta$ are universally quantified over logical
variables.  The {\sc L:Extend} rule uses the host proof assistant to
require that the triple on $f$'s body is proved for every possible
logical variable $y$.

@@

% \section{Polynomial Potential}

% \paragraph{Index Sets}

% Let $V$ be a set of variables.  An \emph{index} $I \in \ind(V)$ is a
% family that maps two-element sets of variables to natural numbers,
% that is,
% $$
% I = (i_{\{x,y\}})_{\{x,y\} \subseteq V} \; .
% $$
% %
% We identify a family $I$ with the set $\{ (\{x,y\},i_{\{x,y\}})
% \mid \{x,y\} \subseteq V\}$.

% Let $\ind(V)$ denote the set of all such indices.  We write $\ind$
% instead of $\ind(V)$ if the set of variables $V$ is fixed or obvious
% from the context.
% %
% We assume that allways $0 \in V$ and sometimes write $i_x$ instead of $i_{\{x,0\}}$.

% The \emph{degree} $\deg(I)$ of an index $I = (i_{\{x,y\}})_{\{x,y\}
%   \subseteq V}$ is defined as
% $$
% \deg(I) = \sum_{\{x,y\} \in V} i_{\{x,y\}} \;.
% $$
% We define $\ind_k(V) = \{ I \mid I \in \ind(V) \text{ and } \deg(i) \leq k
% \}$ to be the set of indices of degree at most $k$.

% \paragraph{Resource Polynomials}

% Let $V$ be a set of variables.  An index $I \in \ind(V)$ denotes a
% \emph{base polynomial} $P_I : \states \to \N$ for $V$ that maps a
% program state $H$ to product of binomial coefficients (a natural
% number).  We define
% $$
% P_I(H) = \prod_{{\{x,y\}} \subseteq V} \binom {|H(x){-}H(y)|} {i_{\{x,y\}}} \; .
% $$
% %
% A \emph{resource polynomial} $R$ for the variable set $V$ is a
% non-negative linear combination of the base polynomials for $V$.

% \paragraph{Potential Annotations}

% A \emph{potential annotation} for the variable set $V$ is a family
% $$Q = (q_I)_{I \in \ind(V)}$$ 
% of non-negative rational numbers.  Such an annotation denotes the
% resource polynomial $R_Q$ that is defined by
% $$
% R_Q(H) = \sum_{I \in \ind(V)} q_I \cdot P_I(H) \; .
% $$
% %
% We say that the potential annotation $Q$ is of degree $k$ if $q_I = 0$
% for $I \in \ind(V)$ with $\deg(I) > k$.

% \paragraph{Additive Shifts}

% Let $Q$ be a potential annotation for a variable set $V$ and let
% $\{x,y\} \subseteq V$ be a two-element variable set.  The
% \emph{additive shift} with respect to $\{x,y\}$ is a potential
% annotation $\shift_{\{x,y\}}(Q) = (q'_I)_{I \in \ind(V)} $ for $V$
% that is defined through
% $$
% q'_I = q_I + q_{I^{\{x,y\}{+}1}} \; .
% $$
% For an index $I = (i_{\{x,y\}})_{\{x,y\} \subseteq V}$ we use the
% notation $I^{\{x,y\}{+}k}$ to denote the index
% $(i'_{\{x,y\}})_{\{x,y\} \subseteq V}$ such that
% $$
% i'_{\{t,u\}} = \left\{
%   \begin{array}{ll}
%     i_{\{t,u\}} + k  & \text{if } \{t,u\} = \{x,y\} \\
%     i_{\{t,u\}} & \text{otherwise}
%   \end{array}
% \right.
% \;.
% $$
% %
% The additive shift for natural numbers reflects the identity 
% \begin{equation}
% \label{eq:shift}
% \sum_{0 {\leq} i \leq {k}} q_i \binom{n+1}{i} = \sum_{0 {\leq} i \leq {k}} (q_i{+}q_{i+1}) \binom{n}{i}
% \end{equation}
% where $q_{k+1} = 0$.  It is used in the effect system if the
% difference $n+1$ between two variables $x,y$ decreases by one.

% \begin{lemma} Let $V$ be a set of variables with $x,y \in V$ and let
%   $H$ be a program state. Let $|H'(t) {-} H'(u)| = |H(t) {-} H(u)|$
%   for $\{t,u\} \neq \{x,y\}$ and let $|H'(x) {-} H'(y)| = |H(x) {-}
%   H(y)| - 1$.
%   %
%   If $Q' = \shift_{\{x,y\}}(Q)$ then $R_Q(H) = R_{Q'}(H')$.
% \end{lemma}

% We now study the effect of multiple simultaneous shifts.  Let $Q$ be a
% resource annotation for a variable set $V$ and let $U_1,\ldots,U_n
% \subseteq V$ with $|U_i| = 2$ for all $i$ and $U_i \neq U_j$ for $i
% \neq j$ be pairwise distinct two-element variable sets.  The
% simulations additive shift $\shift_{U_1,\ldots,U_n}(Q)$ of $Q$ with
% respect to $U_1,\ldots,U_n$ is defined by
% $$
% \shift_{U_1,\ldots,U_n}(Q) = \shift_{U_1}( \cdots \shift_{U_n}(Q) \cdots ) \; .
% $$
% %
% \begin{proposition}
%   Let $V$ be a set of variables and let $U_1,\ldots,U_n$ be pairwise
%   distinct two-element variable sets.  Let $|H'(x) {-} H'(y)| = |H(x)
%   {-} H(y)|$ for $\{x,y\} \not\in \{U_1,\ldots,U_n\}$ and let $|H'(x)
%   {-} H'(y)| = |H(x) {-} H(y)| - 1$ for $\{x,y\} \in
%   \{U_1,\ldots,U_n\}$.
%   %
%   If $Q' = \shift_{U_1,\ldots,U_n}(Q)$ then $R_Q(H) = R_{Q'}(H')$.
% \end{proposition}
% %
% As shown by the following lemma, the order in which the shifts for the
% individual $U_i$ are applied is insignificant.
% %
% \begin{lemma}
%   Let $\sigma : \{1,\ldots,n\} \to \{1,\ldots,n\}$ be a
%   permutation. Then $\shift_{U_1,\ldots,U_n}(Q) =
%   \shift_{U_{\sigma(1)},\ldots,U_{\sigma(n)}}(Q)$.
% \end{lemma}
% %
% For reasons of efficiency in the constraint generation, we give a more
% direct formula for the simultaneous shift.  Let $I \in \ind(V)$ and
% let $U_1,\ldots,U_n$ be pairwise distinct two-element variable sets.
% We define the index $I^{U_1,\ldots,U_n + k}$ as the family $(i'_{\{x,y\}})_{\{x,y\} \subseteq V}$ such that
% $$
% i'_{\{t,u\}} = \left\{
%   \begin{array}{ll}
%     i_{\{t,u\}} + k  & \text{if } \{t,u\} \in \{U_1,\ldots,U_n\} \\
%     i_{\{t,u\}} & \text{otherwise}
%   \end{array}
% \right.  \;.
% $$

% %
% \begin{lemma}
%   Let $V$ be a variable set and let $U_1,\ldots,U_n$ be pairwise
%   distinct two-element variable sets.
%   %
%   Let $Q = (q_I)_{I \in \ind(V)}$ be a resource annotation for
%   $V$ and let $ Q' = (q'_I)_{I \in \ind(V)}$ where
%   $$
%   q'_I = \sum_{\{j_1,\ldots,j_m\} \subseteq \{1,\ldots,n\} } q_{I^{U_{j_1},\ldots,U_{j_m}+1}} \; .
%   $$
%   Then $Q' = \shift_{U_1,\ldots,U_n}(Q)$.
% \end{lemma}

@@
